---
title: "PicnicHealth Hemophilia B subset Analysis"
author: "Jaison Jacob"
date: "2022-10-23"
output: html_document
---

```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(ggplot2)
library(PicnicHealth)
library(PicnicBlanket)
library(magrittr)
library(DT)
library(ggpmisc)
library(kableExtra)

filename <- "/Users/poops/Projects/hemB/sample_dataset/data"
hemB <- load_data_set(filename)
```
```{r other_settings, echo = FALSE}
blanks <- rep(c(' ', '\n'),5) # This will print five blank lines between plots. 

AddBreak <- function() {
for (i in blanks)
  cat(i)
}
```


## Hemophilia B test dataset exploration

The goal of this analysis is to assess the hypothesis that compared to mild or moderate hemophilia B patients, patients with severe hemophilia B have:

1. higher healthcare utilization 
2. more bleeds.

Our definition of the phenotypic severity will be as follows:

* Patients with a baseline factor IX level of <1% have severe hemophilia
* Patients with a baseline factor IX level of 1-5% have moderate hemophilia
* Patients with a baseline factor IX level of 5-50% have mild hemophilia 


Table 1. Patient characteristics for this exploratory cohort is shown above.
```{r table_1, include = TRUE}
table_one( hemB)

```


Now let us start diving into the ***<span style="color:red">baseline</span>*** data and create data subsets for easy exploration and analysis.  The code below parses the baseline dataset and creates a severity index using our definitions above based on the Coagulation factor IX activity actual/normal in Platelet poor plasma by Coagulation assay.

```{r clean_baseline_data, include = TRUE}
severity_data <- hemB$baseline %>% 
  dplyr::filter(measurement_concept_id == "8b961efa-cc89-55c3-bcd8-e83c74cb4612") %>% 
  dplyr::mutate(value_as_range = tidyr::replace_na(value_as_range,"{}")) %>% 
  dplyr::mutate(tidyjson::spread_all(value_as_range)) %>% 
  dplyr::mutate(baseline_value = dplyr::case_when(
    recorded_value_type == "number" ~ value_as_number, 
    type %in% c("lessThan", "lessThanOrEqualTo") ~ borderValue - 0.1, 
    type %in% c("greaterThanOrEqualTo", "greaterThan") ~ borderValue + 0.1, 
    type == "between" ~ (lowValue + highValue)/2)) %>% 
  dplyr::mutate(b_severity = dplyr::case_when(
    baseline_value <  1 ~ "Severe", 
    baseline_value >= 1 & baseline_value <= 5 ~ "Moderate", 
    baseline_value > 5 & baseline_value <=  50 ~ "Mild", 
    baseline_value > 50 ~ "Normal")) %>% 
  dplyr::group_by(person_id) %>% 
  dplyr::filter(baseline_value == min(baseline_value)) %>%
  dplyr::distinct( person_id, baseline_value, b_severity ) %>%
  dplyr::arrange( baseline_value )

severity_data$person_id = factor(severity_data$person_id, 
                                 levels = severity_data$person_id)
```

```{r show_severity, echo = FALSE}
severity_data %>%
  kbl(caption = "Table 2. Severity Data") %>%
    kable_styling(bootstrap_options = "striped", full_width = F, position = "center", 
                  html_font = "Cambria")
```



Now we will explore the ***<span style="color:red">visit</span>*** data to obtain healthcare utilization information from *Inpatient Visits* and *Outpatient Visits* columns.

```{r hosp_data, include = TRUE, echo = TRUE}
hospitalization_data <- hemB$visit %>%
  dplyr::mutate(hosp_days = difftime(hemB$visit$visit_end_date,
                                     hemB$visit$visit_start_date, units = "days") %>%
                  as.integer) %>% 
  dplyr::group_by( person_id, visit_concept_name) %>%
  dplyr::mutate( visit_type_count = dplyr::n(),
                 total_hosp_days = sum(hosp_days)) %>%
  dplyr::ungroup() %>%
  dplyr::distinct( person_id, visit_concept_name, visit_type_count, total_hosp_days) %>%
  tidyr::pivot_wider( names_from = visit_concept_name,
                      values_from = c(visit_type_count, total_hosp_days)) %>%
  dplyr::select( person_id,
                 in_visits  = `visit_type_count_Inpatient Visit`,
                 out_visits = `visit_type_count_Outpatient Visit`,
                 overnight_hosp = `total_hosp_days_Inpatient Visit`) %>%
  dplyr::left_join( severity_data, by = "person_id") %>%
  dplyr::arrange( baseline_value ) %>%
  dplyr::mutate( total_visits = tidyr::replace_na(in_visits,0) + 
                   tidyr::replace_na(out_visits, 0))

hospitalization_data$person_id = factor(hospitalization_data$person_id, 
                                 levels = hospitalization_data$person_id)

age_data <- hemB$person %>%
  dplyr::mutate(age_at_enroll = difftime(enrollment_date, 
                                         date_of_birth,
                                         units = "days")/365 %>%
  as.numeric()) %>%
  dplyr::select(person_id,age_at_enroll, 
                sex = sex_concept_name, 
                race  = race_concept_name) %>%
  dplyr::left_join( severity_data, by = "person_id") %>%
  dplyr::arrange( baseline_value ) %>%
  dplyr::mutate(age_at_enroll = sprintf("%0.1f", age_at_enroll))

age_data$person_id = factor(age_data$person_id, 
                                 levels = age_data$person_id)

age_data <- dplyr::left_join( age_data, severity_data)

hospitalization_data <- dplyr::left_join( age_data, hospitalization_data) %>%
  dplyr::mutate( label1 = sprintf("Age = %s; Overnights = %s", 
                                 age_at_enroll,in_visits),
                 label2 = sprintf("Age = %s; Overnights = %s", 
                                 age_at_enroll,out_visits),
                 label3 = sprintf("Age = %s; Overnights = %s", 
                                 age_at_enroll,overnight_hosp),
                 )

```


```{r show_hosp, echo = FALSE}
hospitalization_data %>%
  dplyr::select(-c(label1, label2, label3)) %>%
  kbl(caption = "Table 2. Hospitalization Data") %>%
    kable_styling(bootstrap_options = "striped", full_width = F, position = "center", 
                  html_font = "Cambria")
```

```{r plot_hosp, echo = FALSE, warning=FALSE}

ggplot( hospitalization_data,
        aes( x= in_visits, 
             y=person_id, 
             label = label1,
             fill = b_severity)) +
    geom_bar(  stat = "identity") +
    geom_text( aes(x=10), size = 2, hjust = 0) +
    ggtitle("Figure 1. In-Patient Visits vs Patient ID") +
  ylab( "Patient ID") +
  xlab("In-patient Visits")+
  labs( fill = "Severity")

cat("\n\n")
cat("\n\n")

ggplot( hospitalization_data,
        aes( x= out_visits, 
             y=person_id, 
             label = label2,
             fill = b_severity)) +
    geom_bar(  stat = "identity") +
    geom_text( aes(x=10), size = 2, hjust = 0) +
  ggtitle("Figure 2. Out-Patient Visits vs Patient ID") +
  ylab( "Patient ID") +
  xlab("Out-patient Visits") +
  labs( fill = "Severity")

cat("\n\n")
cat("\n\n")

ggplot( hospitalization_data,
        aes( x= overnight_hosp, 
             y=person_id, 
             label = label3,
             fill = b_severity)) +
    geom_bar(  stat = "identity") +
    geom_text( aes(x=10), size = 2, hjust = 0) +
  ggtitle("Figure 3. Overnight Hospitalization vs Patient ID") +
  ylab( "Patient ID") +
  xlab("Overnight Hospitalization")+
  labs( fill = "Severity")

```

No strong correlations seen between severeity as defined and either In-patient or Out-patient vist counts.  However, there seems to be a weak correlation between overnight hospitalizations.  There seems to be one patient classified as moderate that seems to be breaking the trend.  

Next, let us look at the ***<span style="color:red">bleed_rate</span>*** to see if there are any straightforward correlations between severity as we have defined vs number of bleeds. 

```{r bleed_rate, include = TRUE}
bleed_data <- hemB$bleed_rate %>%
  dplyr::mutate( bleed_days = 
                   dplyr::case_when(
                     bleed_window_type == "Bleeds between two dates" ~
                       (bleed_window_end_date -
                          bleed_window_start_date) %>% as.numeric,
                     bleed_window_type == "Bleeds per unit time" ~
                       dplyr::case_when( bleed_window_unit_time == "Per 6 Months" ~ 6*30.5 ,
                                         bleed_window_unit_time == "Per Month" ~ 30.5,
                                         bleed_window_unit_time == "Per Year" ~ 365,
                                         bleed_window_unit_time == "Per 6 Weeks" ~ 6*7,
                                         bleed_window_unit_time == "Per 2 Weeks" ~ 2*7,
                                         bleed_window_unit_time == "Per Week" ~ 7,
                                         bleed_window_unit_time == "Per 2 Months" ~ 2*30.5,
                                         bleed_window_unit_time == "Per 3 Months" ~ 3*30.5
                       )  
                   )
  ) %>%
  dplyr::mutate(bleed_count_as_range = tidyr::replace_na(bleed_count_as_range,"{}")) %>% 
  dplyr::mutate(tidyjson::spread_all(bleed_count_as_range)) %>% 
  dplyr::mutate(bleed_count_in_period = dplyr::case_when(
    recorded_bleed_count_type == "number" ~ bleed_count_as_number, 
    type %in% c("lessThan") ~ borderValue - 1, 
    type == "between" ~ (lowValue + highValue)/2)) %>%
  dplyr::left_join(age_data)
  
bleed_data$person_id = factor( bleed_data$person_id,
                               levels = levels(age_data$person_id))

bleed_data <- bleed_data %>% droplevels %>%
  dplyr::mutate( std_bleed_cound = bleed_count_as_number/bleed_days*365)

```

Since the number of bleeds will depend on the number of days where this information is recorded, we should ideally be looking at some standardized varibale.  We define a new variable ___standardized bleed count = 365 * (total number of bleeds per patient)/(total number of bleed days)___.
```{r plot_bleed_rate, echo = FALSE, warning=FALSE}

ggplot( bleed_data, aes(x=bleed_window_collection_date, 
                        y= reorder(person_id,std_bleed_cound) ,
                        size=std_bleed_cound, 
                        color = b_severity)) + 
  geom_point() +
  ggtitle("Figure 4. Patients ordered by standardized bleed count vs bleed date") +
  ylab( "Patient ordered by standardized bleed count") +
  xlab("Overnight Hospitalization")+
  labs( fill = "Severity")

cat("\n\n\n\n")

ggplot( bleed_data, aes(x=bleed_window_collection_date, 
                        y= person_id , 
                        size=bleed_count_as_number,
                        color = b_severity)) +
  geom_point() +
  ggtitle("Figure 5. Bleed count at every bleed window collection data vs Patient ID") +
  ylab( "Patient ID") +
  xlab("Overnight Hospitalization")+
  labs( fill = "Severity")

cat("\n\n\n\n")

```

```{r bleed_data_2, include = TRUE}

bleed_data %>%
  dplyr::group_by( person_id) %>%
  dplyr::mutate( total_bleed_days = sum(bleed_days),
                 total_bleeds = sum(bleed_count_as_number),
                 overall_bleed_rate = total_bleeds/total_bleed_days*365 ) %>%
  dplyr::ungroup() %>%
  dplyr::distinct( person_id, age_at_enroll, sex, race,total_bleed_days,total_bleeds )

```
